Analyzing tweets to detect social media events

### 2020-05-06 Rmd/jse-2019-ms.Rmd
minor reword
updated ms

---
title: 'Detecting social media events through tweet content analysis'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

@@

While our tweet collection enables us to address many possible research questions, the 
content of tweets over time particularly intrigued us. We 
hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we'd detect social media events through changes in tweet 
topic content over time. We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via latent Dirichlet allocation modeling of collections of tweets and 3) reflections on 
using this data set in research mentoring of undergraduate 
students.

@@

We sought to validate our hypothesis that we could detect a major social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
analyze tweets before, during, and after the National Football League's 
2015 Super Bowl. We fitted latent Dirichlet allocation models for each of five 
distinct one-hour periods. The first period began approximately 48h 
before the Super Bowl halftime show. Subsequent time periods started at -24, 
0, +24, and +48 hours after the approximate start time of the halftime show.

We defined each time period to be a single corpus of tweets. Each tweet constituted a document. We then fitted latent Dirichlet allocation models to each of the five corpora.

@@

>>>>

---
title: 'Analyzing tweets to detect social media events'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

@@

While our tweet collection enables us to address many possible research questions, the 
dynamic content of tweets over time particularly intrigued us. We 
hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we'd detect social media events through changes in tweet 
topic content over time. We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via latent Dirichlet allocation modeling of collections of tweets and 3) reflections on 
using this data set in research mentoring of undergraduate students.

@@

@@

We sought to validate our hypothesis that we could detect a major social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
analyze tweets before, during, and after the National Football League's 
2015 Super Bowl. We fitted latent Dirichlet allocation models for each of five 
distinct one-hour periods. The first period began approximately 48h 
before the Super Bowl halftime show. Subsequent time periods started at -24, 
0, +24, and +48 hours after the approximate start time of the halftime show. 
We defined each time period to be a single collection, or corpus, of tweets. Each tweet 
constituted a "document", in the terminology of text analysis. We then fitted latent 
Dirichlet allocation models to each corpus.

We used several criteria to evaluate latent Dirichlet allocation model fits, with 
emphasis on choosing a reasonable number of topics per corpus.
Our strategy involved both visualization and more quantitative approaches to model 
evaluation. For every model, we created one word cloud per topic.

We then inspected topic contents at each of the five time points.
